---
title: "Preprocessing Raw Data, Base R Version"
author: "Keith Baggerly"
date: "`r Sys.Date()`"
output: github_document
---

# Outline

We want to organize the three raw datasets (Potti, NCI60, and GEO)
in roughly consistent fashion to allow for analysis and cross-checking. 
Here, we define this to mean producing a gene by sample expression matrix
for each dataset (with rownames and colnames), and data frame of sample
information for each. Some challenges for each dataset include

* parsing the header row of the Potti dataset to extract information 
about drug (e.g., Adria, Doce, etc), contrast group (0 or 1), and 
column index within the overall matrix for reference. 
* arranging the table of information from the NCI into matrix form
with columns using the cell line name from cell name and the replicate
array suffix from ID, while also looking for and removing duplicate
row entries (since there are more rows than the 180 arrays we suspect
were used would generate)
* combining the results from the sensitive and resistant cohorts, 
and correcting the mislabeling of one sample in the annotation. 

We save the data and annotation matrices in results, one RData
file per dataset. 

# Libraries

```{r load_libs, message=FALSE}

library(here)
library(readr)
library(GEOquery)
library(tidyr)
library(magrittr)

```

# The Potti Data

First, we load the expression data into a matrix.

```{r load_potti, warning=FALSE}

potti_tibble <- 
  read_delim(here("data", "chemo.zip"), delim = "\t")

potti <- as.matrix(potti_tibble[, 2:ncol(potti_tibble)])
rownames(potti) <- potti_tibble[[1]]

rm(potti_tibble)

potti[1:5, 1:5]

```

Next, we parse the column names to extract sample information.

```{r parse_column_names}

## Strip the suffixes R added to duplicate names on loading

raw_column_names <- 
  gsub("_[[:digit:]]+$", "", colnames(potti))

## The last character is either 0 or 1

contrast_group <- 
  substr(raw_column_names, 
         nchar(raw_column_names), nchar(raw_column_names))

## Fill in the gaps, since names are given just
## for the first and last entries for a drug

drug_name <- rep(NA, ncol(potti))
drug_name[nchar(raw_column_names) > 1] <- 
  substr(raw_column_names[nchar(raw_column_names) > 1],
         1, nchar(raw_column_names)[nchar(raw_column_names) > 1] - 1)
for(i in 2:length(drug_name)){
  if(is.na(drug_name[i])){
    drug_name[i] <- drug_name[i - 1]
  }
}

## Recast the column index as a 0-padded string of width 3

raw_index <- sprintf("%03d", c(1:ncol(potti)))

## Join all components together for a more useful name

full_names <- paste(drug_name, 
                    contrast_group, 
                    raw_index,
                    sep = "_")

## Combine into a data frame

potti_info <- 
  data.frame(drug_name = drug_name,
             contrast_group = contrast_group,
             raw_index = raw_index,
             row.names = full_names,
             stringsAsFactors = FALSE)

## Replace the initial (less helpful) column names

colnames(potti) <- full_names

## Clean up

rm(i, drug_name, contrast_group, raw_index, full_names)

## Save the results

save(potti, potti_info,
     file = here("results", "potti_data.RData"))

```

# The NCI60 Data

First, we load the data. 

```{r load_nci60}

nci60_tibble <- 
  read_csv(here("data", "WEB_DATA_NOVARTIS_ALL.zip"))


id_width <- nchar(nci60_tibble$ID[[1]])
all(nchar(nci60_tibble$ID) == nchar(nci60_tibble$ID[[1]]))

```

Next, we look for duplicate entries. From our previous explorations, 
we know there are results for _about_ 180 arrays, assuming
12625 probesets per array.

```{r check_for_dups}

length(unique(nci60_tibble$`Probe Set Name`))
nrow(nci60_tibble) / 180
temp_tab <- table(nci60_tibble$`Probe Set Name`)
temp_tab[1:5]
which(temp_tab != 180)

```

Looking more closely, we see that the number of rows is 12626 * 180, 
so there are precisely 180 more rows than we expect. Checking the
frequencies of the probesets, we see there's just one which is
present more than 180 times: "100\_g\_at", which may well be the 
first probeset in alphabetical order. This is present exactly 360
times, so we suspect the values were repeated for some reason. 

```{r check_odd_probeset}

temp_tib <- nci60_tibble[nci60_tibble$`Probe Set Name` == "100_g_at", ]
temp_tib[1:3,]
temp_tib[181:183,]
all.equal(temp_tib[1:180, c(1, 3:10)], temp_tib[181:360, c(1, 3:10)])

```

For some reason, this probeset was matched to two gene cluster (GC)
ids, so the values were repeated twice. Since we don't care that 
much about the gene cluster, we can simply drop the rows associated
with the (arbitrarily chosen) latter of the two, GC32977. 



First, we split the ID column, which is a composite of gene associated
information (the GC) and sample associated information (the replicate)
into separate columns.

```{r separate_id_into_pieces}

nci60_tibble <- 
  nci60_tibble %>%
  separate(ID, into = c("gene_id", "replicate"), sep = "_")

nci60_tibble[1:3, ]

```

Then we trim off the unwanted rows.

```{r trim_data}

nci60_tibble <- nci60_tibble[nci60_tibble$gene_id != "GC32977", ]

```

The resulting
table is now properly dimensioned to be reorganized in matrix form. 

Of course, before performing such a reorganization we'd like to check 
whether the sample results are always presented in the same order within
a probeset, because if they are (which we suspect) we can organize things 
faster. The Signal, Detection, and P Value column entries may differ, 
but we expect the rest of the values (annotation) to be consistent. 

Now we check whether all of the sample info is consistently ordered. 

```{r check_sample_orders}

n_samples <- 180
n_genes <- 12625

sample_cols <- c("replicate", "panelnbr", "cellnbr", "pname", "cellname")
for(i in 1:length(sample_cols)){
  temp <- nci60_tibble[[sample_cols[i]]]
  if( all( temp == rep(temp[1:n_samples], n_genes) ) ){
    cat(sample_cols[i], "matches\n")
  }else{
    cat(sample_cols[i], "doesn't match\n")
  }
}

```

Now we check whether all of the gene info is consistently ordered. 

```{r check_gene_orders}

n_samples <- 180
n_genes <- 12625

gene_cols <- c("Probe Set Name", "gene_id", "GENE")
for(i in 1:length(gene_cols)){
  temp <- nci60_tibble[[gene_cols[i]]]
  temp[is.na(temp)] <- ""
  if( all( temp == 
           rep(temp[seq(from = 1, to = n_genes * n_samples, 
                        by = n_samples)], each = n_samples) ) ){
    cat(gene_cols[i], "matches\n")
  }else{
    cat(gene_cols[i], "doesn't match\n")
  }
}

```

Ok, everything matches. Let's reorganize what remains into 
gene information, sample information, and expression values, 
discarding the Detection and P Value columns.

```{r reorg_nci}

key_gene_rows <- seq(1, n_genes * n_samples, n_samples)

nci60_gene_info <- 
  as.data.frame(nci60_tibble[key_gene_rows, c("gene_id", "GENE")])
names(nci60_gene_info) <- c("gene_cluster", "gene_name")
rownames(nci60_gene_info) <- nci60_tibble[["Probe Set Name"]][key_gene_rows]

nci60_sample_info <- 
  as.data.frame(
    nci60_tibble[1:n_samples, 
                 c("cellname", "replicate", "pname",
                   "panelnbr", "cellnbr")])
rownames(nci60_sample_info) <- 
  paste(nci60_sample_info[["cellname"]],
        nci60_sample_info[["replicate"]], sep = "_")

nci60 <- 
  matrix(nci60_tibble[["Signal"]], 
         nrow = n_genes, ncol = n_samples, byrow = TRUE)
rownames(nci60) <- rownames(nci60_gene_info)
colnames(nci60) <- rownames(nci60_sample_info)

```

Now we clean up and save the results. 

```{r save_nci60}

rm(nci60_tibble, temp_tib, temp_tab,
   key_gene_rows, n_genes, n_samples, i, 
   temp, gene_cols, sample_cols)

save(nci60, nci60_gene_info, nci60_sample_info,
     file = here("results", "nci60_data.RData"))

```

# The GEO Data

Here we simply load in the matrices of expression values from
the SOFT files of the two patient cohorts, combine them if their
row names match, and record which samples are resistant and which
are sensitive. 

We load GSE349 (the resistant samples and one mislabeled
sensitive sample) first.

```{r load_gse349, warning=FALSE, message=FALSE}

gse349_gq <-
  getGEO(filename = here("data", "GSE349_family.soft.gz"))

n_genes <- nrow(Table(gse349_gq@gsms[[1]]))
n_samples <- length(gse349_gq@gsms)

gse349 <- matrix(0, nrow = n_genes, ncol = n_samples)
rownames(gse349) <- Table(gse349_gq@gsms[[1]])[, "ID_REF"]
colnames(gse349) <- names(gse349_gq@gsms)

for(i in 1:n_samples){
  if(all(rownames(gse349) == Table(gse349_gq@gsms[[i]])[, "ID_REF"])){
    gse349[, i] <- Table(gse349_gq@gsms[[i]])[, "VALUE"]
  }
}

gse349[1,]

```

Now we repeat the process for GSE350 (the sensitive samples).

```{r load_gse350, warning=FALSE, message=FALSE}

gse350_gq <-
  getGEO(filename = here("data", "GSE350_family.soft.gz"))

n_genes <- nrow(Table(gse350_gq@gsms[[1]]))
n_samples <- length(gse350_gq@gsms)

gse350 <- matrix(0, nrow = n_genes, ncol = n_samples)
rownames(gse350) <- Table(gse350_gq@gsms[[1]])[, "ID_REF"]
colnames(gse350) <- names(gse350_gq@gsms)

for(i in 1:n_samples){
  if(all(rownames(gse350) == Table(gse350_gq@gsms[[i]])[, "ID_REF"])){
    gse350[, i] <- Table(gse350_gq@gsms[[i]])[, "VALUE"]
  }
}

gse350[1,]

```

Check whether the tables can be safely combined...

```{r check_geo_combo}

all(rownames(gse349) == rownames(gse350))

```

Now we bundle the matrices and record the response status. 

```{r combine_geo}

geo <- cbind(gse349, gse350)

geo_sample_info <- 
  data.frame(sample_status = 
               c(rep("Resistant", ncol(gse349)),
                 rep("Sensitive", ncol(gse350))),
             geo_dataset = 
                c(rep("GSE349", ncol(gse349)),
                  rep("GSE350", ncol(gse350))),
             row.names = colnames(geo),
             stringsAsFactors = FALSE)
geo_sample_info["GSM4913", "sample_status"] <- "Sensitive"

```

Now we reorder the columns matrix columns to put all of the resistant 
samples first and all of the sensitive samples next, sorting by GSM
id within a cohort.

```{r reorder_geo}

sorted_geo_names <- 
  c(sort(rownames(geo_sample_info)[
    geo_sample_info$sample_status == "Resistant"]),
    sort(rownames(geo_sample_info)[
    geo_sample_info$sample_status == "Sensitive"]))

geo_sample_info <- 
  geo_sample_info[sorted_geo_names, ]

geo <- geo[, sorted_geo_names]

```

Now we clean up and save the results. 

```{r save_geo}

rm(gse349_gq, gse349, gse350_gq, gse350, i, 
   n_genes, n_samples, sorted_geo_names)

save(geo, geo_sample_info,
     file = here("results", "geo_data.RData"))

```


