---
title: "Check Sample Matches and Correlations"
author: "Keith Baggerly"
date: "`r Sys.Date()`"
output: github_document
---

# Overview

Given what we've seen of the Potti et al and NCI60 datasets, 
we want to see whether we can nail down the precise mappings
used. 

We load the data matrices for Potti et al, the NCI60, and the
docetaxel patient cohort posted to GEO. We check for exact
matches between expression values in the Potti et al and NCI60
datasets, and between the Potti et al and GEO datasets if
the former fails. We also examine the pairwise sample 
correlations between columns to see whether that clarifies
any of the structure present.

Precise matching on data values lets us match all of the 
Potti et al columns for 5 of the 7 drugs to columns in the 
NCI60 dataset; we know which cell lines are being used 
in each of the contrast groups. We find no matches for 
columns associated with Docetaxel (Doce) and Cytoxan (Cytox); correlation
plots show the contents of these columns are wholly uncorrelated
with those in the rest of the data matrix. Examining minimums
shows the minimum values in these columns are (a) all the same, 
and (b) the same as the minimum values we saw in earlier
examination of the GEO data. Again looking for exact matches
shows the Doce and Cytox columns match columns
from the GEO datasets after allowing for an offset of 67 rows
to exclude the Affymetrix control probes reported in GEO. 
These matches are exact for the first 12535 of 12558 probesets
reported; the last 23 use a slightly different order. Importantly, 
the values in the Potti table match the GEO values 
_in the order they are presented at GEO_, but these probeset ids
do not match those used in the Potti et al table, effectively
producing a random scrambling of the data values. 


# Librares and Data

## Libraries

```{r load_libs}

library(here)
library(lattice)

```

## Data

```{r load_data}

load(here("results", "potti_data.RData"))
load(here("results", "nci60_data.RData"))
load(here("results", "geo_data.RData"))

```

# Matching the Cell Lines Used

Our first question is whether we can match the data values supplied
by Potti et al to the NCI60 data, and thus infer the identities of 
the cell lines used to characterize response for each drug. Based 
on our earlier description of the raw data tables, we think checking
for matches of the first one or two probeset values might suffice. 

```{r check_potti_for_nci60_matches}

potti_nci60_names <- rep(NA, ncol(potti))
for(i1 in 1:ncol(potti)){
  temp <- which(nci60["36460_at", ] == potti["36460_at", i1])
  if(length(temp) == 1){
    potti_nci60_names[i1] <- colnames(nci60)[temp]
  }
}

drug_names <- unique(potti_info$drug_name)
contrast_levels <- c(0, 1)
for(i1 in 1:length(drug_names)){
  for(i2 in 1:length(contrast_levels)){
    cat("drug: ", drug_names[i1], " level ", contrast_levels[i2],
        "\n", 
        potti_nci60_names[ 
          (potti_info$drug_name == drug_names[i1]) & 
          (potti_info$contrast_group == contrast_levels[i2])], "\n",
        fill = TRUE)
  }
}

```

We get perfect matches for all cell lines used for 5 of the 7 drugs. 
We get no matches at all for two of drugs: Doce (Docetaxel) and
Cytox (Cytoxan). Failing for all cell lines for these drugs suggests
something more involved might be going on there. 

All of the matches we do see are for the "A" set of cell line replicates,
suggesting the way they dealt with having triplicate measurements of 
each line was to simply take the first in each instance. 

# Check Sample Correlations in the Potti et al Data

```{r get_potti_cors}

potti_cors <- cor(potti)

```

Let's try plotting the pairwise correlations for all columns
in the Potti et al dataset.

```{r plot_raw_potti_cors}

levelplot(potti_cors,
          xlab = "", ylab = "",
          scales = list(alternating=0, tck=0),
          main = "Pairwise Sample Correlations\nin Potti et al Data")

```

The columns for the two drugs (Docetaxel and Cytoxan) stand out clearly. 
They're effectively uncorrelated with the columns of the NCI60 data 
matrix. 

We can also look for sample reuse by checking where the pairwise correlations
are quite high (say above 0.99).

```{r plot_potti_high_cors}

levelplot(potti_cors > 0.99,
          xlab = "", ylab = "",
          scales = list(alternating=0, tck=0),
          main = "Pairwise Sample Correlations > 0.99\nin Potti et al Data")

```

As it happens, there are a few off-diagonal contiguous stretches of
"hits", and these are coming from the Docetaxel and Cytoxan columns.
We can see this by looking at the values for the first probeset.

```{r check_doce_cytox_asymmetry}

## Doce 0, Cytox 1

potti["36460_at", (potti_info$drug_name == "Doce") &
        (potti_info$contrast_group == 0)]
potti["36460_at", (potti_info$drug_name == "Cytox") &
        (potti_info$contrast_group == 1)]

## Doce 1, Cytox 0

potti["36460_at", (potti_info$drug_name == "Doce") &
        (potti_info$contrast_group == 0)]
potti["36460_at", (potti_info$drug_name == "Cytox") &
        (potti_info$contrast_group == 1)]

```

The Docetaxel and Cytoxan columns are identical but for the 
reversal of the contrast group labels. This makes little
sense biologically. 

# Check Other Summaries for Potti et al Data

Given the discrepancy, we also checked whether other summary
statistics applied to the Potti et al data might clarify
the situation. We looked at values typically returned 
by `summary` (most not shown), of which the minimum was
most enlightening. 

```{r plot_potti_minimums}

plot(apply(potti, 2, min),
     ylab = "Minimum Expression Value",
     xlab = "Raw Column Index",
     main = "Minimum Expression for Potti et al Data")

drug_boundaries <- 
  which(potti_info$drug_name[1:(ncol(potti) - 1)] != 
          potti_info$drug_name[2:ncol(potti)])
abline(v = drug_boundaries + 0.5)

drug_centers <- 
  (c(1, drug_boundaries+1) + c(drug_boundaries, ncol(potti))) / 2
drug_center_names <- 
  potti_info$drug_name[floor(drug_centers)]
mtext(drug_center_names, side = 3, 
      at = drug_centers)

```

The minimums for the Doce and Cytox samples all look to be about 
the same. Let's look at the actual values. 

```{r check_min_values}

apply(potti[, potti_info$drug_name == "Doce"], 2, min)
apply(potti[, potti_info$drug_name == "Cytox"], 2, min)

```

All of the minimum values are 5.89822. We've seen this value before. 
This is the minimum value for all of the samples in the datasets
from GEO, which are not cell lines, but rather patient samples.
Since this was one of the cohorts used to test their predictor 
for docetaxel, this may be the test data as opposed to the training
data. It's not at all clear why these should be present for cytoxan, 
nor is it clear why the correlations beteen these samples and the
cell lines. would all be near zero. 

# Matching Potti to GEO

We can explore the hypothesis that some of the Potti data 
might be coming from the GEO data by looking for matching
values for a few entries. We'll start with the first 
probeset in the Potti data, 36460\_at. 

```{r check_first_match}

potti["36460_at", "Doce_0_023"]

which(geo == potti["36460_at", "Doce_0_023"], arr.ind = TRUE)

which(geo == potti["36460_at", "Doce_0_024"], arr.ind = TRUE)

which(geo == potti["36460_at", "Doce_0_025"], arr.ind = TRUE)

```

It appears row 1 of the Potti et al columns is mapping to row 68
of the GEO data. While row 68 of the GEO data is the first
non-control probe row, the probeset id is _not_ the same; the
probeset ordering is different between the NCI60 data and the
GEO data. 

We can check this further by looking at the next few rows
for some of the samples that match.

```{r check_first_few_rows}

colnames(geo)[14]
potti[1:5, "Doce_0_023"]
geo[67 + c(1:5), "GSM4903"]

```

The values match exactly, but since the probesets are
in a different order, the values have essentially been randomly
scrambled with respect to each other. Random scrambling would 
explain why we correlations near zero. We can check how extensive
the offset is for this sample (for which there are 12558 data rows
in the Potti et al table):

```{r check_match_extent}

sum(potti[, "Doce_0_023"] == geo[67 + c(1:nrow(potti)), "GSM4903"])
which(potti[, "Doce_0_023"] != geo[67 + c(1:nrow(potti)), "GSM4903"])

all( sort(potti[12534:nrow(potti), "Doce_0_023"]) == 
       sort(geo[67 + c(12534:nrow(potti)), "GSM4903"]) )

```

We match all but the last 23 of the 12558 entries using the offset, 
and the remaining entries involve the same values, just scrambled
in a different way. This is where this column in the Potti data
comes from. 

Now, there are just 20 columns in the Potti data labeled as belonging
to Docetaxel (and Cytoxan), but there are 24 columns of data in the 
GEO dataset. 

We're simply going to try brute force matching here. 

```{r apply_brute_force}

n_doce <- 20
n_geo <- 24

match_count_matrix <- 
  matrix(0, nrow = n_doce, ncol = n_geo)
rownames(match_count_matrix) <- 
  colnames(potti)[potti_info$drug_name == "Doce"]
colnames(match_count_matrix) <- 
  colnames(geo)

for(i1 in 1:n_doce){
  for(i2 in 1:n_geo){
    match_count_matrix[i1, i2] <- 
      sum(potti[, rownames(match_count_matrix)[i1]] ==
            geo[67 + c(1:nrow(potti)), colnames(match_count_matrix)[i2]])
  }
}

table(match_count_matrix[match_count_matrix > 1000])

```

We have matches of the same extent for all 20 of the Doce columns.  
Let's eyeball roughly where these are. 

```{r plot_potti_doce_geo_matches}

levelplot(match_count_matrix,
          main = "N Matches of Potti Doce Columns\nand GEO Data Columns")

```

These are essentially in linear order, so the sensitive and resistant
patient groups map pretty well to the 0/1 contrast group labels. 

Let's collect all of the name matches into a vector. We can 
add these for Cytoxan too, given that the Doce and Cytox
values are the same except for the contrast group label. 

```{r match_potti_to_geo}

potti_geo_matches <- rep(NA, ncol(potti))
names(potti_geo_matches) <- colnames(potti)
for(i1 in 1:nrow(match_count_matrix)){
  potti_geo_matches[rownames(match_count_matrix)[i1]] <-
    colnames(match_count_matrix)[which(match_count_matrix[i1, ] > 1000)]
}

potti_geo_matches[(potti_info$drug_name == "Cytox") &
                    (potti_info$contrast_group == 0)] <-
  potti_geo_matches[(potti_info$drug_name == "Doce") &
                      (potti_info$contrast_group == 1)]

potti_geo_matches[(potti_info$drug_name == "Cytox") &
                    (potti_info$contrast_group == 1)] <-
  potti_geo_matches[(potti_info$drug_name == "Doce") &
                      (potti_info$contrast_group == 0)]

```

Similarly, let's grab a vector defining the sample status
for the GEO samples used.

```{r record_geo_status}

geo_status <- rep(NA, ncol(potti))
geo_status[!is.na(potti_geo_matches)] <- 
  geo_sample_info[potti_geo_matches[!is.na(potti_geo_matches)],
                  "sample_status"]

```

# Combine all Match Info Into a Data Frame

We now know where all of the columns in the original 
data matrix came from. Here, we combine this into a 
single data frame for later use. 

```{r collate_match_info}

potti_matches <- potti_info
potti_matches$nci60_match <- potti_nci60_names
potti_matches$geo_match <- potti_geo_matches
potti_matches$geo_status <- geo_status

```

# Save the Results

Now we save the results for later use, both as an 
RData object and as a csv file. 

```{r save_matches}

save(potti_matches,
     file = here("results", "potti_matches.RData"))
write.csv(potti_matches,
          file = here("results", "potti_matches.csv"))

```

